Human Factors Influencing Visual Statistical Inference

### 2013-06-20 sociology_paper.tex
impact of graduate degree plot and discussion

{\bf Keywords: \sf statistical graphics, non-parametric test, cognitive phycology, data visualization, exploratory data analysis, data mining, visual analytics} 
\end {abstract}

@@

\begin{figure}[hbtp] 
   \centering
   \includegraphics[width=0.99\textwidth]{electoral-5-13.pdf} 
   \caption{ \label{fig:elect-1} Which one of the plots is the most different from the others?}
\end{figure}

@@

\cline{3-6} \cline{8-11} 
Factor & Level& Est & SE & Zval & $p$-value && Est & SE & Zval & $p$-value  \\ 
  \hline
\multicolumn{2}{l}{Fixed Effect} &  &  &  &  &  & & && \\ 
&$\mu$ & 3.360 & 0.013 & 249.21 & $<$0.001 &   & -0.683 & 0.071 & -9.64 & $<$0.001 \\ 
\hline
Age&18-25&0.000   &---  &---  &---  &  &--- &--- &---& ---\\   
 & 26-30 & 0.058 & 0.013 & 4.50 & $<$0.001 &   & 0.062 & 0.049 & 1.27 & 0.2058 \\ 
  &31-35 & 0.068 & 0.014 & 4.72 & $<$0.001 &   & 0.115 & 0.055 & 2.08 & 0.0376 \\ 
  &36-40 & 0.231 & 0.016 & 14.05 & $<$0.001 &   & 0.310 & 0.063 & 4.93 & $<$0.001 \\ 
  &41-45 & 0.176 & 0.021 & 8.56 & $<$0.001 &   & 0.158 & 0.081 & 1.96 & 0.0501 \\ 
  &46-50 & 0.272 & 0.024 & 11.29 & $<$0.001 &   & 0.141 & 0.096 & 1.47 & 0.1425 \\ 
  &above 50 & 0.352 & 0.018 & 19.19 & $<$0.001 &   & 0.147 & 0.071 & 2.06 & 0.0394 \\ 
  \hline
Country& United States&0.000   &---  &---  &---  &  &--- &--- &---& ---\\  
  &India & 0.101 & 0.011 & 9.11 & $<$0.001 &   & 0.058 & 0.043 & 1.33 & 0.183 \\ 
  &Rest of world & -0.129 & 0.009 & -13.82 & $<$0.001 &   & 0.185 & 0.035 & 5.22 & $<$0.001 \\ 
  \hline
Education &High school or less&0.000   &---  &---  &---  &  &--- &--- &---& ---\\     
 &Under grad courses & 0.042 & 0.013 & 3.25 & 0.0011 &   & -0.083 & 0.050 & -1.65 & 0.098 \\ 
  &Under grad degree & -0.037 & 0.012 & -3.21 & 0.0013 &   & -0.044 & 0.045 & -0.97 & 0.3307 \\ 
  &Graduate courses & 0.117 & 0.013 & 9.12 & $<$0.001 &   & 0.070 & 0.050 & 1.42 & 0.1566 \\ 
  &Graduate degree & 0.046 & 0.011 & 4.12 & $<$0.001 &   & 0.182 & 0.043 & 4.22 & $<$0.001 \\ 
  \hline
Gender & Female&0.000   &---  &---  &---  &  &--- &--- &---& ---\\      
  &Male & 0.078 & 0.009 & 8.26 & $<$0.001 &   & 0.055 & 0.036 & 1.5 & 0.1331 \\ 
  \hline
\multicolumn{2}{l}{Random Effect} &  &  &  &  &  & & && \\ 
& lineup & 0.082 & 0.287 &  &  &   & 5.259 & 2.293 &  &  \\ 
 & Error & 0.479 & 0.692 &  &  & & && &\\ 
   \hline
\end{tabular}
}
\label{tbl:model_result_demographics}
\end{table}

@@

The graduate degree holders are significantly different as well. The positive parameter estimate indicates that they perform better and the probability of correct response is higher than other education levels. As expected, there is no significant difference between male and female performances.

@@

Effect of graduate degree is estimated as 0.182 which is highly significant. For an 18-25 year old female in the United States, the expected difference in the probability of correct response, if she has a graduate degree as compared to a high school degree, is, for a lineup with random effect of zero:
$$ \frac {\exp(-0.683+0.182)}{1+\exp(-0.683+0.182)}-  \frac {\exp(-0.683)}{1+\exp(-0.683)} = 0.377- 0.336 =0.041.$$ 
For a lineup with a predicted random effect of one standard deviation, i.e. we move away one standard deviation of lineup variation by adding 2.293 to each of the exponent terms in the equation, this difference reduces to 0.024. For two standard deviations it further reduces to 0.003 which is practically very small. Figure~\ref{fig:practical_impact_demographics} illustrates this example showing how the change in probability diminishes for larger variability of lineup difficulty. A similar calculation shows the same negligible impact of age level 36-40 (0.0533 at most)  and country (0.0424 at most) on the probability of correct response. Thus even though some of the factor levels are significant, their practical impact is negligible because of high variability of lineup difficulties.

\begin{figure}[htbp] 
   \centering
%   \includegraphics[width=6.5in]{learning_trend.pdf} 
    \includegraphics[width=3in]{practical_impact_demographics.pdf} 
   \caption{ Change in probability of correct response due to graduate degree as compared to high school degree diminishes rapidly as the variability in lineup difficulty $\sigma_{\ell}$ increases.}
   \label{fig:practical_impact_demographics}
\end{figure}

@@

It is found that observer demographics have significant influence on performance in time taken to evaluate a lineup. Some variations among the factors are observed in terms of probability of correct evaluation. Age group 36-40, Countries other than India and United states, People who have a graduate degree are significantly different. But their practical impact on probability to correctly evaluate a lineup is very small and in some cases very negligible as the lineup difficulties increase or decreases. Gender does not have any effect on performance.

Individual learning trend is observed in both time taken and observer performance. Some individuals improved the performances while others showed decrease in their performances over attempts. But the overall performance of the observers does not increase through successive attempts while evaluating multiple lineups. This result is very important as it suggests that the power estimated for visual inference using human subject experiment is robust and may not change if those participants are allowed to give feedback again. The skill in evaluating the lineup in shorter time gets improved over successive evaluations. The earlier evaluations take significantly longer than the later evaluations.  This investigation suggest that the skilled person may only do it faster.

@@

The subjects participating this study may not necessarily know about statistical graphics. The numerous pilot studies done with more trained participants suggest that power of visual statical inference may be higher for observers knowledgeable about statistical graphics. The fact that a person with graduate degree performs better may be associated with this. But a graduate degree does not necessarily mean to have training on statistical graphics. More experiments may be needed to learn about the differences in power with trained and a non-trained observer in terms of knowledge about statistical graphics.

@@

%\begin{table}[hbtp]
%\caption{Amazon mechanical turk experiments and their properties. Duration in hours per 100 tasks show the popularity of some tasks compared to others.}
%\centering
%\begin{tabular}{rlrrrrrrr}
%  \hline
%& Experiment& \multicolumn{2}{c}{ Total Task}& Average & \multicolumn{2}{c}{Duration (hour)} & Payment & Pay rate\\
%\cline{3-4} \cline{6-7}
%Serial & description & submitted & rejected & time(min) & Actual & 100 task& \$/task & \$/hour\\ 
%  \hline
%1 & Boxplot & 406 & 106 & 10.68 & 146.48 & 36.08 & 0.50 & 2.81 \\ 
%  2 & Scatterplot & 359 &   9 & 10.80 & 42.68 & 11.89 & 1.00 & 5.58 \\ 
%  3 & Contaminated plot & 219 &  19 & 13.53 & 126.17 & 57.61 & 1.00 & 2.22 \\ 
%  4 & Polar vs Cartesian & 110 &  10 & 20.65 & 11.65 & 10.59 & 1.00 & 2.91 \\ 
%  5 & Hist vs density & 234 &  37 & 17.85 & 41.57 & 17.76 & 1.00 & 3.36 \\ 
%  6 & Violin vs boxplot & 417 &  17 & 17.95 & 105.87 & 25.39 & 1.00 & 3.34 \\ 
%  7 & Group separation & 106 &   6 & 16.13 & 5.15 & 4.86 & 1.00 & 3.72 \\ 
%  8 & Sine Illusion & 101 &   1 & 16.52 & 78.38 & 77.60 & 1.00 & 3.63 \\ 
%  9 & Gene expression & 103 &   3 & 12.47 & 11.27 & 10.94 & 0.50 & 2.41 \\ 
%  10 & Test normality & 406 &   6 & 22.70 & 74.35 & 18.31 & 1.00 & 2.64 \\ 
%   \hline
%\end{tabular}
%\label{tbl:mturk}
%\end{table}
%
%
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=3in]{rejected_task.pdf}
%      \includegraphics[width=3in]{task_duration.pdf} 
%   \caption{Percentage of rejected tasks and duration of each experiment in hour per 100 tasks for each of the 10 experiments. Most of the tasks got rejected for box plot experiment.  Even though the sine illusion experiment took longest to finish the rejection rate is lowest for this experiment.}
%   \label{fig:task_duration}
%\end{figure}
\clearpage
\section{Some Plots of Exploratory Data Analysis}

@@

\clearpage
\section{Electoral Building Lineups and Results}
%Figures~\ref{fig:elect-1} and \ref{fig:elect.2} through~\ref{fig:elect.5}

>>>>

{\bf Keywords: \sf statistical graphics, non-parametric test, cognitive phycology, data visualization, exploratory data analysis, data mining, visual analytics.} 
\end {abstract}

@@

\begin{figure}[hbtp] 
   \centering
   \includegraphics[width=0.99\textwidth]{electoral-5-13.pdf} 
%      \includegraphics[width=0.99\textwidth]{electoral-1-1.pdf} 
   \caption{ \label{fig:elect-1} Which one of the plots is the most different from the others?}
\end{figure}

@@

\cline{3-6} \cline{8-11} 
Factor & Level& Est & SE & Zval & $p$-value && Est & SE & Zval & $p$-value  \\ 
  \hline
\multicolumn{2}{l}{Fixed Effect} &  &  &  &  &  & & && \\ 
&$\mu$ & 3.360 & 0.013 & 249.21 & $<$0.001 &   & -0.683 & 0.071 & -9.64 & $<$0.001 \\ 
\hline
Age ($\alpha$)& 18-25&0.000   &---  &---  &---  &  &--- &--- &---& ---\\   
 &26-30 & 0.058 & 0.013 & 4.50 & $<$0.001 &   & 0.062 & 0.049 & 1.27 & 0.2058 \\ 
  &31-35 & 0.068 & 0.014 & 4.72 & $<$0.001 &   & 0.115 & 0.055 & 2.08 & 0.0376 \\ 
  &36-40 & 0.231 & 0.016 & 14.05 & $<$0.001 &   & 0.310 & 0.063 & 4.93 & $<$0.001 \\ 
  &41-45 & 0.176 & 0.021 & 8.56 & $<$0.001 &   & 0.158 & 0.081 & 1.96 & 0.0501 \\ 
  &46-50 & 0.272 & 0.024 & 11.29 & $<$0.001 &   & 0.141 & 0.096 & 1.47 & 0.1425 \\ 
  &above 50 & 0.352 & 0.018 & 19.19 & $<$0.001 &   & 0.147 & 0.071 & 2.06 & 0.0394 \\ 
  \hline
Country($\kappa$)& United States&0.000   &---  &---  &---  &  &--- &--- &---& ---\\  
  &India & 0.101 & 0.011 & 9.11 & $<$0.001 &   & 0.058 & 0.043 & 1.33 & 0.183 \\ 
  &Rest of world & -0.129 & 0.009 & -13.82 & $<$0.001 &   & 0.185 & 0.035 & 5.22 & $<$0.001 \\ 
  \hline
Education($\tau$) &High school or less&0.000   &---  &---  &---  &  &--- &--- &---& ---\\     
 &Under grad courses & 0.042 & 0.013 & 3.25 & 0.0011 &   & -0.083 & 0.050 & -1.65 & 0.098 \\ 
  &Under grad degree & -0.037 & 0.012 & -3.21 & 0.0013 &   & -0.044 & 0.045 & -0.97 & 0.3307 \\ 
  &Graduate courses & 0.117 & 0.013 & 9.12 & $<$0.001 &   & 0.070 & 0.050 & 1.42 & 0.1566 \\ 
  &Graduate degree & 0.046 & 0.011 & 4.12 & $<$0.001 &   & 0.182 & 0.043 & 4.22 & $<$0.001 \\ 
  \hline
Gender ($\gamma$)& Female&0.000   &---  &---  &---  &  &--- &--- &---& ---\\      
  &Male & 0.078 & 0.009 & 8.26 & $<$0.001 &   & 0.055 & 0.036 & 1.50 & 0.1331 \\ 
  \hline
\multicolumn{2}{l}{Random Effect} &  &  &  &  &  & & && \\ 
& lineup($\sigma_{\ell}$) & 0.082 & 0.287 &  &  &   & 5.259 & 2.293 &  &  \\ 
 & Error($\sigma$) & 0.479 & 0.692 &  &  & & && &\\ 
   \hline
\end{tabular}
}
\label{tbl:model_result_demographics}
\end{table}

@@

The graduate degree holders are significantly different as well. The positive parameter estimate indicates that they perform better and the probability of correct response is higher than other education levels. There is no significant difference between male and female performances.

@@

Effect of graduate degree is estimated as 0.182 which is highly significant. For an 18-25 year old female in the United States, the expected difference in the probability of correct response, if she has a graduate degree as compared to a high school degree, is, for a lineup with random effect of zero:
$$ \frac {\exp(-0.683+0.182)}{1+\exp(-0.683+0.182)}-  \frac {\exp(-0.683)}{1+\exp(-0.683)} = 0.377- 0.336 =0.041.$$ 
For a lineup with a predicted random effect of one standard deviation, i.e. we move away one standard deviation of lineup variation by adding 2.293 to each of the exponent terms in the equation, this difference reduces to 0.024. For two standard deviations it further reduces to 0.003 which is practically very small. Figure~\ref{fig:practical_impact_graduate} illustrates this example showing the proportions of correct responses due to graduate degree as compared to high school degree. The highest practical difference in proportions of correct response is estimated as 0.045 which is very negligible. The difference further diminishes if we move one or two standard deviation of lineup variability. A similar calculation shows the same negligible impact of age level 36-40 (0.0533 at most)  and country (0.0424 at most) on the probability of correct response. Thus even though some of the factor levels are statistically significant, their practical impact is negligible.

\begin{figure}[htbp] 
   \centering
%   \includegraphics[width=6.5in]{learning_trend.pdf} 
    \includegraphics[width=4in]{practical_impact_graduate.pdf} 
   \caption{Proportion of correct responses due to graduate degree as compared to high school degree for an 18-25 year old female in the United States. Even though graduate degree is statistically significant, the largest practical difference in proportion correct is  0.045 which is very negligible. The difference even diminishes as we move away one or two standard deviations ($\sigma_{\ell} =2.293$) of lineup variability.}
   \label{fig:practical_impact_graduate}
\end{figure}

%\begin{figure}[htbp] 
%   \centering
%%   \includegraphics[width=6.5in]{learning_trend.pdf} 
%    \includegraphics[width=3in]{practical_impact_demographics.pdf} 
%   \caption{ Change in probability of correct response due to graduate degree as compared to high school degree diminishes rapidly as the variability in lineup difficulty $\sigma_{\ell}$ increases.}
%   \label{fig:practical_impact_demographics}
%\end{figure}

@@

Human demographics have significant influence on performance in time taken to evaluate a lineup. Some variations among the factors are observed in terms of probability of correct evaluation. Age group 36-40, Countries other than India and United states, People who have a graduate degree are significantly different. But their practical impact on probability to correctly evaluate a lineup is very negligible and in some cases it diminishes  as the lineup difficulties increase or decrease. Gender does not have any significant effect on performance. Thus there may be differences in time taken to evaluate a lineup for different human demographics but the practical impact of demographics on the performance is very negligible. This result is very important for the power of visual test as it demonstrates the robustness of the test for different human factors.

Individual learning trend is observed in both time taken and observer performance. Some individuals improved the performances while others showed decrease in their performances over attempts. But the overall performance of the observers does not increase through successive attempts while evaluating multiple lineups. This result suggests that the power estimated for visual inference using human subject experiment is robust and may not change if those participants are allowed to give feedback again. The skill in evaluating the lineup in shorter time gets improved over successive evaluations. The earlier evaluations take significantly longer time than the later evaluations.  It suggests that the skilled person may only do it faster.

@@

The subjects participating this study may not necessarily know about statistical graphics. The numerous pilot studies done with more trained participants suggest that power of visual statical inference may be higher for observers who have advanced knowledge about statistical graphics. The fact that a person with graduate degree performs better may be associated with this. But a graduate degree does not necessarily mean to have training on statistical graphics. More experiments may be needed to learn about the differences in power with trained and a non-trained observer in terms of knowledge about statistical graphics.

@@

%\begin{table}[hbtp]
%\caption{Amazon mechanical turk experiments and their properties. Duration in hours per 100 tasks show the popularity of some tasks compared to others.}
%\centering
%\begin{tabular}{rlrrrrrrr}
%  \hline
%& Experiment& \multicolumn{2}{c}{ Total Task}& Average & \multicolumn{2}{c}{Duration (hour)} & Payment & Pay rate\\
%\cline{3-4} \cline{6-7}
%Serial & description & submitted & rejected & time(min) & Actual & 100 task& \$/task & \$/hour\\ 
%  \hline
%1 & Boxplot & 406 & 106 & 10.68 & 146.48 & 36.08 & 0.50 & 2.81 \\ 
%  2 & Scatterplot & 359 &   9 & 10.80 & 42.68 & 11.89 & 1.00 & 5.58 \\ 
%  3 & Contaminated plot & 219 &  19 & 13.53 & 126.17 & 57.61 & 1.00 & 2.22 \\ 
%  4 & Polar vs Cartesian & 110 &  10 & 20.65 & 11.65 & 10.59 & 1.00 & 2.91 \\ 
%  5 & Hist vs density & 234 &  37 & 17.85 & 41.57 & 17.76 & 1.00 & 3.36 \\ 
%  6 & Violin vs boxplot & 417 &  17 & 17.95 & 105.87 & 25.39 & 1.00 & 3.34 \\ 
%  7 & Group separation & 106 &   6 & 16.13 & 5.15 & 4.86 & 1.00 & 3.72 \\ 
%  8 & Sine Illusion & 101 &   1 & 16.52 & 78.38 & 77.60 & 1.00 & 3.63 \\ 
%  9 & Gene expression & 103 &   3 & 12.47 & 11.27 & 10.94 & 0.50 & 2.41 \\ 
%  10 & Test normality & 406 &   6 & 22.70 & 74.35 & 18.31 & 1.00 & 2.64 \\ 
%   \hline
%\end{tabular}
%\label{tbl:mturk}
%\end{table}
%
%
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=3in]{rejected_task.pdf}
%      \includegraphics[width=3in]{task_duration.pdf} 
%   \caption{Percentage of rejected tasks and duration of each experiment in hour per 100 tasks for each of the 10 experiments. Most of the tasks got rejected for box plot experiment.  Even though the sine illusion experiment took longest to finish the rejection rate is lowest for this experiment.}
%   \label{fig:task_duration}
%\end{figure}

%\clearpage
\section{Electoral Building Lineups and Results}
%Figures~\ref{fig:elect-1} and \ref{fig:elect.2} through~\ref{fig:elect.5}

%\clearpage
\section{Some Plots of Exploratory Data Analysis}

@@
